A Tale of Two Mechanisms: How mediators act and interact in dynamic capabilities theory\\ Lock, Stock and Two Convoking Mechanisms: Towards a holistic theory of dynamic capabilitie

### 2019-11-18 article4_jbr.rnw
Paper 4 ready for submission to Tommy

@@

\begin{frontmatter}

\title{Dynamic capabilities and competitive advantage}
%\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

@@

\begin{abstract}
TBD

@@

\begin{keyword}
\texttt{Dynamic Capabilities, Deliberate Learning, Dynamics }
\end{keyword}

@@

<<loading, results="asis", echo=FALSE >>=
setwd("/Users/larshovdanmolden/Documents/git/article4")
library(xtable)
library(semPLS)
load("./data/resart4.RData")
r <- main$resmat
rm  <- main$resmatmed
l  <- main$loadings
ic <- data.frame(i[,c(1,4)]);colnames(ic) <- c("b","t")
ic$rn  <- 1:nrow(ic)
bb <- main$models$both
bm <- main$models$mmed
br <- main$models$rmed
#foolars=0.99
@

@@

Extant literature on dynamic capabilities are increasingly focusing on how firm resources
are changed by dynamic capabilities as a response to changes in the environment
\citep{Schilke2018,Eriksson2014}, thus placing firm resources as a mediator between
dynamic capabilities and competetive advantage \citep{Helfat2007}. Less focus has been
placed on the nature of these underlying firm resources (ref) and how they are capturing a
rather heterogeneous bulk of routines, capabilities, assets and capital. Hence, there is a
need to disentangle the black box of firm resources to provide a more nuanced
understanding of different types of firm resources and how they interact to shape firm
competitive advantage \cite{Schilke2018}. Indeed, this gap in the literature is explicitly
emphasized by \cite{Schilke2018} where they 'see an interesting opportunity for future
work to add greater richness to our understanding of the mechanism of resource-base
change, given this mechanism’s central role in many foundational works (..) and the
diverse ways in which resource changes can potentially come about' (p. 419).

The conventional mechanism explored in the extant literature is related to changes in the
firm`s resource base \citep{Eriksson2014,Schilke2018}. This mechanism is also included in
the dominating definition of dynamic capabilities found in \citep{Helfat2007}: 'A dynamic
capability is the capacity of an organization to purposefully create, extend or modify its
resource base' (p. 4). And although this definition opens up for a variety of components
included in the term 'resource base', a point we discuss later in this paper, most of the
subsequent contributions have focused on changes in operational routines (ref),
capabilities (ref), practices (ref) or tangible assets (ref). This focus, albeit very
important to the development of dynamic capabilities literature, seems to leave out a
considerable part of a firm`s 'resource base', namely those resources that are tied to how
the employees of the firm sees their mission, internalize information and effectuates
change. These traits of the organization is very much related to concepts of
'organizational culture' (ref), 'employee cognition' (ref)' 'organizational genes'
\citep{Nelson1982} and 'sense-making' \citep{Weick1995} to name but a few. However, in
relations with dynamic capabilities one promising avenue for exploring mechanisms beyond
the conventional ones described above, makes a clear distinction between \emph{behavioral}
and \emph{non-behavioral} objects of change \citep{Verona2011}. These non-behavioral
objects of change contain psychological objects such as cognitive frames, motivation and
identity that are important for strategy formation and implementation (Gavetti etc, vince
and gabriel ch 15 of book by verona) as they 'concerns how decisions or actions are
shaped, subverted and transformed by emotions, and (..) determine the "way we do things
here"' (Vince and gabriel). In this respect they work as firm resources in generating
outcomes such as performance and competitive advantage. This notion is perhaps expressed
most clearly by Peter Drucker`s famous quote about how 'culture eats strategy for lunch'
(ref) insofar as. Moreover, these non-behavioral objects of change are themselves far
from static and are found to change over time as responses to external change
\citep{Zollo2016} thus making them a natural part of the mechanisms underpinning how
dynamic capabilities work \citep{Schilke2018,Verona2011}. Still, far less is know about
the non-behavioral objects of change and how they work as a mechanism in the dynamic
capabilities theory.

@@

To shed light on this important question we theorize on the role of a particular type of
non-behavioral object of change, \emph{strategic cognition}, and how it acts as a
mediator between dynamic capabilities and firm competitive advantage and hence has a
pivotal place in the literature. We build on important work on the concept of
organizational cognition (ref), non-behavioral object of change (ref) and cognitive
psychology (kahneman and tversky),
and hypothesize about how strategic cognition acts as a mediator in the same way as
conventional behavioral objects of change (such as routines and capabilities) and that
these two are complementary to a certain degree. Consequently, we argue, dynamic capabilities takes
the function of a 'conductor' by orchestrating routines and capabilities, and the role of
a 'coach' in stimulating strategic cognition. We use the case of Microsoft's transition
into the cloud computing business as an illuminating case for demonstrating how the roles
of 'conductor' and 'coach' coexist and act partly complimentary in generating strategic
change and competitive advantage.

SOMETHING ABOUT COMP ADV / why interesting outcome holy grail helfat schilke

This paper makes the following contributions:

The rest of this paper is organized as follows:

@@

When Satya Nadella took over as CEO of Microsoft in March 2014, he took over a company in
big trouble. The share price was just north of 30 USD and the company had just launched a
7.2 billion USD acquisition of the failing Nokia company. The story about how Nadella
turned Microsoft around to a valuation of 105 USD pr share and with successful mergers
along the way is a story of excellent 'resource orchestration' as a response to exercising
dynamic capabilites, specifically through sensing, seizing and transforming resources to
adapt to changing market conditions \citep{Teece2007}. Nadella and his firm did a
considerable reorientation towards the cloud and reconfigured firm capabilities to fit
this objective - almost a school example of exercising dynamic capabilities.

Moreover, and to the point of this paper, through a series of talks, interviews and his
own book Nadella put considerable emphasis on how Microsoft managed to change their
'mindset', that is their 'cognitive frames' (Nadella 2017): 'I focused on what would be
our grandest endeavor, the biggest hurdle - transforming the Microsoft culture'
(p. 89). Drawing on the concept of a 'growth mindset' \citep{Dweck2016} Nadella sees
culture as a 'complex system of individual mindsets' and that it encompasses how 'an
organization thinks and act' (ibid p.90).

Satya Nadella is very much on the same page as recent contributions on how cognitive
frames of an organization (i.e. its non-behavioral object of action) is a pivotal part in
how firm resources shape outcomes such as performance and competitive advantage
\citep{Barney1991a,Barney2001}, and how it consequently plays a role in how dynamic
capabilities work along these lines \cite{Verona2011}. Microsoft under Nadella did
transform the company by changing the behavioral objects of action (i.e. capabilities and
routines), but also through changing the non-behavioral objects of action (i.e. the
strategic cognition). In other words, the story of Microsoft`s transformation suggest that
dynamic capabilities worked through mediating factors to influence competetive advantage,
but that these mediating factors exist as two distinctly different concepts working in
tandem. Figure \ref{fig:fig1} presents a conceptual depiction of the paths running through
these two concepts of \emph{capabilities} and \emph{strategic cognition} objects of action
\citep{Zollo2016}. We use a familiar term for the conventional understanding of dynamic
capabilities mechanism as 'orchestration' (ref) as a collective term to describe the
behavioral objects of action (ref). In this process dynamic capabilities play the role of
a 'conductor' orchestrating capabilities of the firm, an understanding that is in line
with a range of the most common definitions of dynamic capabilities \citep{Schilke2018}.

The non-behavioral objects of action has no similar agreed upon analogy, so for the
purpose of our analysis we envision this mechanism to work more as a cognitive process
enabling the organization to tear away from their current understanding and ways in which
things are done (Vince and Gabriel). Whereas the analogy of a 'conductor' suggests a
function that composes an outcome based on the available capabilities, the notion of a
'coach' refers to the function of enabling others to achieve their objectives. The process
is all about changing an organization`s thinking from the automated and well-known, to the
more cognitively aware, moving from what Kahneman and Tversky coins the automated 'system
one' to the deliberate 'system two' thinking (ref). We will now turn to the two functions
of dynamic capabilities in turn.

@@

Satya Nadella set out to transform Microsoft by investing heavily in artificial
intelligence and cloud computing capabilities. This entailed hiring new people, upgrading
the skills and abilities of the current staff, and acquire infrastructure and
hardware. Additionally, Microsoft was able to acquire and integrate a number of companies
to support the strategic direction the company was on. Companies such as GitHub, LinkedIn
and Citrus data became important platforms for developing new capabilities within the
company (Nadella).

This role of 'conductor' is the conventional way of understanding dynamic capabilities. It
stems from one of the most common definitions \citep{Helfat2007} as well as the analogy of
the process as 'resource orchestration' \citep{Sirmon2011}. It is also found in the
empirical literature handling various functional forms of dynamic capabilities such as
acquisition targeting \citep{Bingham2015}, alliancing \cite{Schilke2014}, new product
development \cite{Danneels2008}, innovation capabilities \citep{Breznik2014}, marketing
capabilities \cite{Brune2009} and alliance capabilities \cite{Kale2002,kale2007}.

@@

Juxtaposed to the experiences from Microsoft, the case of Polaroid brings another set of
experiences to our discussion on non-behavioral objects of action (see \citep{Tripsas2000}
for a detailed account). Cognitive inertia on the part of the organization played a
pivotal part in the failing of the company`s digital imaging effort
\citep{Verona2011}. In other words, Polaroid did not exercise dynamic capabilities to
change the cognition of the organization. Similar patterns are found in Kodak and Anderson Consulting (ref
Henderson and Kaplan 2005), and Smith Corona (Danneels 2010). These examples highlight the
absence of 'dynamic capabilities specific to the adaptation of cognitive frames, first,
and consequently of operating processes' \citep[p. 541]{Verona2011}. This is also
highlighted at the core of the dynamic capabilities literature: 'Dynamic capabilities are about doing the
right things, at the right time, based on new product (and process) development, unique
managerial orchestration processes, a strong and {\bf change-oriented organizational
  culture}' (our emphasis) \cite[p]{Teece2014}.

When dynamic capabilities are executed it is as a response to changes in the environment
in order to stay competitive. This processes relates to creating strategic change
\citep{Helfat2007} and entails changing strategic direction and objectives of the
organization. This strategic change requires, in addition to modifying, extending and
creating firm capabilities, also some kind of alignment with the organization as a whole -
what is often referred to as 'how we do things here' \citep{Nelson1982}, and how an
organization thinks and act (NAdella). This alignment is possible only if the employees
share a common cognitive frame related to the the strategic direction of the company
enabling them to make sense of the changes ahead. This is what we have coined
\emph{strategic cognition} and has its expression in Microsoft's development of a growth
mindset. Strategic cognition is needed to get employees on board as well as capture the
efficiency of non-directed change. Getting on board refers to the general willingness of
the organization to accept and value the importance of some decision or system 'without
being prompted or required by formal governance
mechanisms'\citep[p. 812]{Simsek2009}. Moreover, When individuals of a firm buys into
changes they will enhance the full organization`s ability to not only act upon changes
(Marchand et.al 2000), but also in 'timely responsiveness' (Kohli et.al 1993,
Simsek2009). Consequently, strategic cognition not only contributes to getting people on
board, it also maintain a more efficient change process where the individuals take more
responsibilities for the change itself and leverages their individual knowledge in the
endeavor. For a firm to create real strategic change as enabled by their dynamic
capabilities and hence stay competitive, is will need to act as a 'coach' to instill a
strategic cognition in the organization.

@@

@@

We contacted all the firms that were registering R\&D activities during May to December
2005 yielding a total of 1721 firms. Using a web-based questionnaire we included items to
measure the key constructs of this paper. Each firm received an email with a link to the
questionnaire within a month after their registration for the R\&D tax credit. This
initial contact was followed by two subsequent email reminders. A total of 1199 (70 \% of
the initial sample size) answered and returned the questionnaire. In the second phase of
the study each of these 1199 were contacted again in 2014, 9 years later, with the lion
share of the firms contacted during the spring and summer of 2014. All received a similar
web-based questionnaire containing the same measures of the focal constructs of this
paper. 283 of the firms returned filled-in questionnaire.

@@

The focal constructs of this paper are dynamic capabilities, operational capabilities,
strategic cognition as measured by proactiveness, and competitive advantage. These are all
operationalized using the survey data described above. Table \ref{tab:item} in the
appendix describes the items used to operationalize the focal constructs and their
internal validity as measured by Dillon-Goldstein Rho which is the preferred metric for
this purpose \citep{Hair2019}. The results show that our key constructs can be represented
by the items in the survey, with a minor caveat. The measure of strategic cognition as
proactiveness loads only partly on one item (item 3). Statsistically, a loading of
$\Sexpr{l[["lcog"]][3]}$ is on the weaker side of traditional 'rules of thumb'
\citep{Hair2010}, but from a theoretical stand point the measure makes sense and is
established in the literature (REF to EO).

@@

Although we recognize at the outset that the concept of dynamic capabilities and their
underlying resource components are very challenging to research in a systematic and
econometric fashion \citep{McKelvie2009}, we follow the argument in the
literature that more empirical work is necessary to test and refine the dynamic
capabilities concept and how it is related to the evolutionary economic theory
\citep{Arend2009,McKelvie2009} \todo{Rosenblom, 2000; Verona and Ravasi, 2003;}.  It
is in this spirit that the research reported in this paper has been undertaken.

@@

The nine year lag between our points of measurement has its distinct advantages with
respect to capturing long term trajectory of the firms. For example, the operating
capabilities at time $t=1$ is very likely a function of operating capabilities at time
$t=0$ as well as other factors such as dynamic capabilities. Indeed, extant literature
argues for this path dependency of resources and routines
\citep{Helfat2007,Winter2003}. These capabilities are said to be serial-correlated and
path dependent. However, as dynamic capabilities and operating capabilities are distinctly
different concepts, this kind of path dependency is also not equally clear in this
regard. We could have firms with very high levels of dynamic capabilities at $t=0$ and
equally a poorer level at time $t=1$. Because of the long time lag a lot of factors could
have influenced this decline, and indeed a rather high number of over firms exhibit this
characteristic of declining dynamic capabilities over time. As we regard dynamic
capabilities as 'the first derivative' \citep{Winter2003} (i.e. the rate of change of the
operating capabilites) a large initial value (at time $t=0$) could over estimate the
expected change of the underlying capability at time $t=1$. Moreover, based on the scale
applied (Likert from 1 through 7) a high initial value leaves nowhere to move but
downwards. This is not to say that those firms are likely to develop poorer operating
routines. Nor is to say that those with low dynamic capabilities at time $t=0$ are
rendered useless in developing capabilities going forward towards $t=1$. Rather it
suggests that a long time span may include unnecessary noise in the model. Thus, we
estimated an average of the dynamic capabilities of the firm between time $t=0$ and $t=1$
as a proxy for a more representative measure of the rate of change. Hence, our construct
of $DC_0$ is actually a measure between times $t=0$ and $t=1$. As a robustness check we
repeated the full analysis using 2005 measure as our $DC_0$. This did not severely alter
the conclusions we draw in this paper, but had some minor effects on the significance of
the path between $DC_0$ and $COG_1$. The full alternative analysis is included in the
appendix.

@@

We add a set of control variables expected to explain parts of the variation in our focal
constructs. From the survey we add firm size, firm age, the level of dynamism facing the
firm. Size is the reported number of employees in the firm at $T_1$, and age is the number of
years since establishment.

To partly remedy the problem of common method bias, we add controls of financial data from
the firm´s profit and loss statements, and balance sheet. Specifically, we add the natural
logarithm of the assets to capture the capital intensity of the firm.

>>>>

%\usetikzlibrary{arrows.meta}

@@

\begin{frontmatter}
%Lock, Stock and Two Smoking Barrels lock stock and two mediating factors
%The Lord of the Rings: The Two Towers
%Bravo Two Zero (1999)
%A tale of two mechanisms
%Eating strategy for breakfast?
%Push, skill and two mediating factors

\title{A Tale of Two Mechanisms: How mediators act and interact in dynamic capabilities
  theory\\ Lock, Stock and Two Convoking Mechanisms: Towards a holistic theory of dynamic
  capabilitie }
%\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

@@

\begin{abstract}
This paper investigates strategic cognition as mediating factor in the dynamic
capabilities-competitive advantage relationship. By including strategic cognition into the
analysis in addition to the conventional mediator (operating capabilities) we argue that a
more holistic approach to understanding the workings of dynamic capabilities emerges. We
introduce two expressions of how dynamic capabilities work, namely as a conductor (the
conventional mechanism) and as a coach (the cognitive mechanism). Using repeated
cross-sectional data from Norwegian technology firms our results indicate support for our
holistic model. This has implications for understanding how dynamic capabilites work, as
well as practical implications for how strategic change is implemented in firms.

@@

\begin{keyword}
\texttt{Dynamic Capabilities, strategic cognition, operating capabilities, mechanisms,
  competitive advantage }
\end{keyword}

@@

<<loading, results="asis", echo=FALSE >>=
setwd("/Users/larshovdanmolden/Documents/git/article4")
library(xtable)
library(semPLS)
#load("./data/resart4.RData")
r <- main$resmat
rm  <- main$resmatmed
l  <- main$loadings
ic <- data.frame(i[,c(1,4)]);colnames(ic) <- c("b","t")
ic$rn  <- 1:nrow(ic)
bb <- main$models$both
bm <- main$models$mmed
br <- main$models$rmed
#foolars=0.99
@

@@

Extant literature on dynamic capabilities are increasingly focusing on how firm resources
are changed by dynamic capabilities as a response to changes in the environment
\citep{Schilke2018,Eriksson2014}, thus placing firm capabilities as a mediator between
dynamic capabilities and competetive advantage \citep{Helfat2007}. This notion of
'orchestration' of firm resources, assets and capabilities has indeed been emphasized in
the extant literature \citep{Sirmon2011,Helfat2015}. Less emphasis has been placed on how
other mechanisms can be work in shaping firm competitive advantage. Indeed, this gap in
the literature is explicitly emphasized by \cite{Schilke2018} where they 'see an
interesting opportunity for future work to add greater richness to our understanding of
the mechanism of resource-base change, given this mechanism’s central role in many
foundational works (..) and the diverse ways in which resource changes can potentially
come about' (p. 419).

The conventional mechanism explored in the extant literature is related to changes in the
firm`s resource base \citep{Eriksson2014,Schilke2018}. This mechanism is also included in
the dominating definition of dynamic capabilities found in \citep{Helfat2007}: 'A dynamic
capability is the capacity of an organization to purposefully create, extend or modify its
resource base' (p. 4). This definition opens up for a variety of components
included in the term 'resource base', a point we discuss later in this paper. We coin this
expression of dynamic capabilities as the {\bf conductor} envisioning a conductor
orchestrating firm capabilities.

This notion of a {\bf conductor} albeit very important to the development of dynamic
capabilities literature, seems to leave out a considerable part of a firm`s 'resource
base', namely those resources that are tied to how the employees of the firm sees their
mission, internalize information and effectuates change. These traits of the organization
is very much related to concepts of 'mindset' \citep{Dweck2016}, 'employee cognition'
\citep{Lakoff1987,Gavetti2012} 'organizational genes' \citep{Nelson1982} and
'sense-making' \citep{Weick1995} to name but a few. However, in relations with dynamic
capabilities, one promising avenue for exploring mechanisms, beyond the conventional ones
described above, makes a clear distinction between \emph{behavioral} and
\emph{non-behavioral} objects of change \citep{Verona2011}. These non-behavioral objects
of change contain psychological objects such as cognitive frames, motivation and identity
that are important for strategy formation and implementation \citep{Vince2011,Gavetti2012}
as they 'concerns how decisions or actions are shaped, subverted and transformed by
emotions, and (..) determine the "way we do things here"' \citep[p. 338]{Vince2011}. In
this respect they work as firm resources in generating outcomes such as performance and
competitive advantage. This notion is popularly perhaps expressed most clearly by Peter
Drucker`s famous quote about how 'culture eats strategy for lunch'. \footnote{The debate
  on whether Drucker actually uttered these words are still in the open, but it was
  attributed to him by Mark Fields in 2006} Moreover, these non-behavioral objects of
change are themselves far from static and are found to change over time as responses to
external change \citep{Zollo2016} thus making them a natural part of the mechanisms
underpinning how dynamic capabilities work \citep{Schilke2018,Verona2011}.  Still, far less
is know about the non-behavioral objects of change and how they work as a mechanism in the
dynamic capabilities theory.We coin {\bf coach} to capture the enabling role dynamic
capabilities plays through \emph{strategic cognition} as
juxtaposed to the more directional role played as {\bf conductor} of \emph{operational capabilities}.

@@

To shed light on this important question we theorize on the role of a particular type of
non-behavioral object of change, \emph{strategic cognition}, and how it acts as a mediator
between dynamic capabilities and firm competitive advantage and hence has a pivotal place
in the literature. We build on important work on the concept of organizational cognition
\citep{Lakoff1987} , non-behavioral object of change \citep{Verona2011} and cognitive
psychology \cite{Tversky1983,Kahneman2011} and hypothesize about how strategic cognition
acts as a mediator in the same way as conventional behavioral objects of change (such as
routines and capabilities) and that these two are complementary to a certain
degree. Consequently, we argue, dynamic capabilities takes the function of a {\bf conductor}
by orchestrating routines and capabilities, and the role of a {\bf coach} in stimulating
strategic cognition. We use the case of Microsoft's transition into the cloud computing
business as an illuminating case for demonstrating how the roles of {\bf conductor} and
{\bf coach} coexist and act partly complimentary in generating strategic change and
competitive advantage.

Centering this paper around the dynamic capabilities-competitive advantage nexus stems
from the origins and promises of the dynamic capabilities theory. Competitive advantage is
considered the 'holy grail' of strategic management \citep{Schilke2018} and is indeed the
focal point of important antecedent theories of dynamic capabilities
\citep{Arndt2018}. Consequently, contributions to push the theory forward should enable a
better understanding of \emph{how} dynamic capabilities work to influence firms'
competitive advantage. In this spirit, our paper makes the following contributions to
extant dynamic capabilities theory. First, we provide clarification on the theoretical
role of non-behavioral objects of change as well as a possible empirical
operationalization through the concept of \emph{proactiveness}. Second, we conceptualize
two different expressions of dynamic capabilities (i.e. {\bf conductor} and {\bf coach})
and theorize on how they work in the dynamic capabilities-competitive advantage nexus. We
also extend this reasoning to include how they relate to each other. Finally, we develop and
test an empirical model suited for analyzing mediating constructs on the dynamic
capabilities-competitive advantage nexus. The time-lag of our data in conjunction with
advances in empirical estimations of multiple simultaneous mediators makes for an
interesting model to push our understanding of important mechanisms in the
theory.

The rest of this paper is organized as follows: First, we present a brief discussion on
the key definitions applied in this paper. We then move on to develop hypothesis of the
focal relationships. Third, we present the data and empirical models used to test our
hypothesis. A discussion follows the results from our model, before we conclude by
offering some limitations to our study as well as future directions for study.

@@

When Satya Nadella took over as CEO of Microsoft in March 2014, he took over a company in
big trouble. The share price was just north of 30 USD and the company had just launched a
7.2 billion USD acquisition of the failing Nokia company. The story about how Nadella
turned Microsoft around to a valuation of 105 USD pr share and with successful mergers
along the way is a story of excellent 'capabilities orchestration' as a response to exercising
dynamic capabilites, specifically through sensing, seizing and transforming resources to
adapt to changing market conditions \citep{Teece2007}. Nadella and his firm did a
considerable reorientation towards the cloud and reconfigured firm capabilities to fit
this objective - almost a school example of exercising dynamic capabilities.

Moreover, and to the point of this paper, through a series of talks, interviews and his
own book Nadella put considerable emphasis on how Microsoft managed to change their
'mindset', that is their 'cognitive frames' \citep{Nadella2017}: 'I focused on what would be
our grandest endeavor, the biggest hurdle - transforming the Microsoft culture'
(p. 89). Drawing on the concept of a 'growth mindset' \citep{Dweck2016} Nadella sees
culture as a 'complex system of individual mindsets' and that it encompasses how 'an
organization thinks and act' (ibid p.90).

Satya Nadella is very much on the same page as recent contributions on how cognitive
frames of an organization (i.e. its non-behavioral object of action) is a pivotal part in
how firm resources shape outcomes such as performance and competitive advantage
\citep{Barney1991a,Barney2001}, and how it consequently plays a role in how dynamic
capabilities work along these lines \cite{Verona2011}. Microsoft under Nadella did
transform the company by changing the behavioral objects of action (i.e. capabilities and
routines), but also through changing the non-behavioral objects of action (i.e. the
strategic cognition). In other words, the story of Microsoft`s transformation suggest that
dynamic capabilities worked through mediating factors to influence competetive advantage,
but that these mediating factors exist as two distinctly different concepts working in
tandem. Figure \ref{fig:fig1} presents a conceptual depiction of the paths running through
these two concepts of \emph{capabilities} and \emph{strategic cognition} objects of action
\citep{Zollo2016}. We use a familiar term for the conventional understanding of dynamic
capabilities mechanism as 'orchestration' as a collective term to describe the
behavioral objects of action \citep{Verona2011}(ref). In this process dynamic capabilities play the role of
a {\bf conductor} orchestrating capabilities of the firm, an understanding that is in line
with a range of the most common definitions of dynamic capabilities \citep{Schilke2018}.

The non-behavioral objects of action has no similar agreed upon analogy, so for the
purpose of our analysis we envision this mechanism to work more as a cognitive process
enabling the organization to tear away from their current understanding and ways in which
things are done \citep{Vince2011}. Whereas the analogy of a {\bf conductor} suggests a
function that composes an outcome based on the available capabilities, the notion of a
{\bf conductor} refers to the function of enabling others to achieve their objectives. The process
is all about changing an organization`s thinking from the automated and well-known, to the
more cognitively aware, moving from what Kahneman and Tversky coins the automated 'system
one' to the deliberate 'system two' thinking \citep{Kahneman2011}. We will now turn to the two functions
of dynamic capabilities in turn.

@@

Satya Nadella set out to transform Microsoft by investing heavily in artificial
intelligence and cloud computing capabilities. This entailed hiring new people, upgrading
the skills and abilities of the current staff, and acquire infrastructure and
hardware. Additionally, Microsoft was able to acquire and integrate a number of companies
to support the strategic direction the company was on. Companies such as GitHub, LinkedIn
and Citrus data became important platforms for developing new capabilities within the
company \cite{Nadella2017}.

This role of {\bf conductor} is the conventional way of understanding dynamic capabilities. It
stems from one of the most common definitions \citep{Helfat2007} as well as the analogy of
the process as 'resource orchestration' \citep{Sirmon2011}. It is also found in the
empirical literature handling various functional forms of dynamic capabilities such as
acquisition targeting \citep{Bingham2015}, alliancing \cite{Schilke2014}, new product
development \cite{Danneels2008}, innovation capabilities \citep{Breznik2014}, marketing
capabilities \cite{Brune2009} and alliance capabilities \cite{Kale2002,kale2007}.

@@

Juxtaposed to the experiences from Microsoft, the case of Polaroid brings another set of
experiences to our discussion on non-behavioral objects of action (see \citep{Tripsas2000}
for a detailed account). Cognitive inertia on the part of the organization played a
pivotal part in the failing of the company`s digital imaging effort \citep{Verona2011}. In
other words, Polaroid did not exercise dynamic capabilities to change the cognition of the
organization. Similar patterns are found in Kodak and Anderson Consulting
\citep{Kaplan2005}, and Smith Corona \citep{Danneels2010}. These examples highlight the
absence of 'dynamic capabilities specific to the adaptation of cognitive frames, first,
and consequently of operating processes' \citep[p. 541]{Verona2011}. This is also
highlighted at the core of the dynamic capabilities literature: 'Dynamic capabilities are
about doing the right things, at the right time, based on new product (and process)
development, unique managerial orchestration processes, a strong and {\bf change-oriented
  organizational culture}' (our emphasis) \cite[p]{Teece2014}.

When dynamic capabilities are executed it is as a response to changes in the environment
in order to stay competitive. This processes relates to creating strategic change
\citep{Helfat2007} and entails changing strategic direction and objectives of the
organization. This strategic change requires, in addition to modifying, extending and
creating firm capabilities, also some kind of alignment with the organization as a whole -
what is often referred to as 'how we do things here' \citep{Nelson1982}, and how an
organization thinks and act \citep{Nadella2017}. This alignment is possible only if the
employees share a common cognitive frame related to the the strategic direction of the
company enabling them to make sense of the changes ahead. This is what we have coined
\emph{strategic cognition} and has its expression in Microsoft's development of a growth
mindset. Strategic cognition is needed to get employees on board as well as capture the
efficiency of non-directed change. Getting on board refers to the general willingness of
the organization to accept and value the importance of some decision or system 'without
being prompted or required by formal governance
mechanisms'\citep[p. 812]{Simsek2009}. Moreover, When individuals of a firm buys into
changes they will enhance the full organization`s ability to not only act upon changes
\citep{Marchand2004}, but also in 'timely responsiveness' \citep{Kohli1993}. Consequently,
strategic cognition not only contributes to getting people on board, it also maintain a
more efficient change process where the individuals take more responsibilities for the
change itself and leverages their individual knowledge in the endeavor. For a firm to
create real strategic change as enabled by their dynamic capabilities and hence stay
competitive, is will need to act as a {\bf coach} to instill a strategic cognition in the
organization.

@@

\subsection{Coaching conductor}

Although Satya Nadella and Microsoft put strong emphasis on the cognitive side of
strategic change (the role as {\bf coach}), acquisition and development of capabilities
were also very much on their mind (the role as {\bf conductor}). It was not about letting
'culture eat strategy for breakfast', but rather serving culture and strategy as
complementary side-dishes. Nadella emphasized 'getting the right team in place', 'build
new and surprising partnerships', 'be ready to catch next wave of innovation and platform
shift' as things Microsoft needed to get right. These meant investing in new people as
well as upgrading the capabilities of the existing staff \cite{Nadella2017}. Meanwhile, he
also stressed that they needed to 'drive cultural change from top to bottom' and 'reframe
our opportunity for a mobile- and cloud-first world'. (ibid p 95-96). In Microsoft, it
clearly seems, it was exactly by being both a {\bf coach} and a {\bf conductor} at the
same time that results would come. In other words, the giant software company strongly
emphasized that the two mechanisms were complementary.

When dynamic capabilities are executed in a firm being able to both adapt the capabilities
and the cognitive frames simultaneously makes sense. A solid strategic cognition in tandem
with relevant and strong capabilities would reinforce each other. This has also been
supported in previous research where firm success has been shown to relate to the
interconnection between resources and capabilities, and culture and attitude
\citep{Verona2003}, and high levels of social capital \citep{Blyler2003}. Moreover, the
lack of one (e.g. strategic cognition) can lead to unfavorable outcomes even though
capabilites and resources are aligned for the needed strategic change
\citep{Tripsas2000}. This leads us to our final hypothesis:

\emph{H3: Strategic cognition and operating capabilities are complementary and thus
  mutually reinforcing their individual effect on competitive advantage}

@@

We contacted all the firms that were registering R\&D activities during May to December
2005 yielding a total of 1721 firms. Using a web-based questionnaire we included items to
measure the key constructs of this paper. Each firm received an email with a link to the
questionnaire within a month after their registration for the R\&D tax credit. This
initial contact was followed by two subsequent email reminders. A total of 1199 (70 \% of
the initial sample size) answered and returned the questionnaire. In the second phase of
the study each of these 1199 were contacted again in 2014, 9 years later, with the lion
share of the firms contacted during the spring and summer of 2014. All received a similar
web-based questionnaire containing the same measures of the focal constructs of this
paper. 262 of the firms returned filled-in questionnaire.

@@

The focal constructs of this paper are dynamic capabilities, operating capabilities,
strategic cognition as measured by proactiveness, and competitive advantage. These are all
operationalized using the survey data described above.  To capture \emph{operating
  capabilites} we look at the definition of these by \citep{Winter2003} as those
capabilities that enables the firm to 'make a living' in the short term, i.e. contributing
to the technical fitness of the firm \citep{Helfat2007}. Thus we chose to survey a firm's
ability to handle core functions of daily operations such as operational routines,
operational management, customer service and marketing. To capture strategic cognition we
apply the definition derived above. Of course such a concept can have many
actual expressions, but in the context of studyin competititve advantage the
organization's attitude towards competition seems like a plausible starting point. In
other word a shared cognitive frame of an overall competitive orientation is useful. Thus,
we adopt the idea of measuring a strategic posture of the firm \citep{Covin1989} and
particularly the notion of \emph{proactiveness} \citep{covin1991,Wales2016}.

Table \ref{tab:item} in the appendix describes the items used to operationalize the focal
constructs and their internal validity as measured by Dillon-Goldstein Rho which is the
preferred metric for this purpose \citep{Hair2019}. The results show that our key
constructs can be represented by the items in the survey, with a minor caveat. The measure
of strategic cognition as proactiveness loads only partly on one item (item
3). Statsistically, a loading of $\Sexpr{l[["lcog"]][3]}$ is on the weaker side of
traditional 'rules of thumb' \citep{Hair2010}, but from a theoretical stand point the
measure makes sense and is established in the literature \cite{covin2012}.

@@

Although we recognize at the outset that the concept of dynamic capabilities and their
underlying resource components are very challenging to research in a systematic and
econometric fashion \citep{McKelvie2009}, we follow the argument in the literature that
more empirical work is necessary to test and refine the dynamic capabilities concept and
how it is related to the evolutionary economic theory \citep{Arend2009,McKelvie2009}.  It
is in this spirit that the research reported in this paper has been undertaken.
% \todo{Rosenblom, 2000; Verona and Ravasi, 2003;}

@@

The nine year lag between our points of measurement has its distinct advantages with
respect to capturing long term trajectory of the firms. For example, the operating
capabilities at time $t=1$ is very likely a function of operating capabilities at time
$t=0$ as well as other factors such as dynamic capabilities. Indeed, extant literature
argues for this path dependency of resources and routines
\citep{Helfat2007,Winter2003}. These capabilities are said to be serial-correlated and
path dependent. However, as dynamic capabilities and operating capabilities are distinctly
different concepts, this kind of path dependency is also not equally clear in this
regard. We could have firms with very high levels of dynamic capabilities at $t=0$ and
equally a poorer level at time $t=1$. Because of the long time lag a lot of factors could
have influenced this decline, and indeed a rather high number of over firms exhibit this
characteristic of declining dynamic capabilities over time. As we regard dynamic
capabilities as 'the first derivative' \citep{Winter2003} (i.e. the rate of change of the
operating capabilites) a large initial value (at time $t=0$) could over estimate the
expected change of the underlying capability at time $t=1$. Moreover, based on the scale
applied (Likert from 1 through 7) a high initial value leaves nowhere to move but
downwards. This is not to say that those firms are likely to develop poorer operating
routines. Nor is to say that those with low dynamic capabilities at time $t=0$ are
rendered useless in developing capabilities going forward towards $t=1$. Rather it
suggests that a long time span may include unnecessary noise in the model. Thus, we
estimated an average of the dynamic capabilities of the firm between time $t=0$ and $t=1$
as a proxy for a more representative measure of the rate of change. Hence, our construct
of $DC_0$ is actually a measure between times $t=0$ and $t=1$. As a robustness check we
repeated the full analysis using 2005 measure as our $DC_0$. This did not severely alter
the conclusions we draw in this paper, but had some minor effects on the significance of
the path between $DC_0$ and $COG_1$.

@@

We add a set of control variables expected to explain parts of the variation in our focal
constructs. From the survey we add firm size and firm age. Size is the reported number of
employees in the firm at $T_1$, and age is the number of years since establishment.

To partly remedy the problem of common method bias, we add controls of financial data from
the firm´s profit and loss statements, and balance sheet as robustness
checks. Specifically, we add the natural logarithm of the assets to capture the capital
intensity of the firm.
